One-arm spirals in locally isothermal, radially structured self-gravitating discs

### 2014-11-15 paper/appendix.tex
simplified discussion

>>>>

\section{Relation between horizontal Lagragian displacements for
  local, low frequency $m=1$ disturbances}\label{horizontal_displacements}
Here, we aim to relate the horizontal Lagrangian displacements $\xi_R$
and $\xi_\phi$ in the local approximation. Using the local solution to
the Poisson equation 
\begin{align}
  \delta \Phi_m = -\frac{2\pi G}{|k|} \delta\Sigma_m 
\end{align}
\citep{shu91}, the linearised azimuthal equation of motion becomes 
\begin{align} 
  - \ii\sbar \delta v_{\phi m}  + \frac{\kappa^2}{2\Omega}\delta v_{Rm} = -\frac{\ii
    m}{R\Sigma}\left(c_s^2 - \frac{2\pi G
      \Sigma}{|k|}\right)\delta\Sigma_m. 
\end{align}
Next, we replace the surface density perturbation
$\delta \Sigma_m = -\ii k \Sigma \xi_R$, and use the expressions
\begin{align}
  &\delta v_{Rm} = -\ii\sbar\xi_R,\\
  &\delta v_{\phi m} = -\ii\sbar\xi_\phi - \frac{\ii R
    \p_R\Omega}{\sbar} \delta v_{Rm}
\end{align}
\citep{papaloizou85} to obtain 
\begin{align}
  -\sbar^2\xi_\phi - 2\ii\sbar\Omega \xi_R =
  \frac{m}{kR}\left(\kappa^2 - \sbar^2\right)\xi_R, 
\end{align}
where the dispersion relation Eq. \ref{dispersion} was used. 
In the local approximation, $|kR|\gg m$ by assumption.%  Furthermore,  
% Also, for low frequency $m=1$ modes $\sbar \simeq -\Omega$, so the
% quantity $|\kappa^2 - \sbar^2| \ll \Omega^2$ in a nearly Keplerian disc
% ($\kappa\sim \Omega$). 
 Hence the RHS of this equation can be neglected. Then 
\begin{align}
  \xi_\phi \simeq -\frac{2\ii\Omega}{\sbar}\xi_R.
\end{align}
For low-frequency $m=1$ modes we have $\sbar \simeq -\Omega$, so
$\xi_\phi\simeq 2\ii\xi_R$, as used in the main text.

### 2014-11-15 paper/discussion.tex
simplified discussion

\subsection{Unstable interaction between low-frequency $m=1$ spirals
  and the background disc due to an imposed temperature gradient}
The $m=1$ spiral that emerges in our simulations have low frequency,
grows slowly, and its spatial form is consistent local theory  
(\S\ref{fargo_m1}). Then to a first approximation, we can regard it 
as a neutral, low-frequency $m=1$ mode in satifying the 
dispersion relation Eq. \ref{dispersion}.  
To demonstrate possible instability, %we 
%first construct a neutral low-frequency mode using local theory, 
we show that the torque density acting on such a mode due to the background
temperature gradient is negative, which enforces the mode, because it
has negative angular mometum.  
% Note that
% we will not be assuming a self-gravitating disc until the end of this
% section.

We begin by recalling the expression for the torque density due to a
locally isothermal equation of state,
\begin{align}
  T_\mathrm{BG} &= -\frac{m}{2}\frac{dc_s^2}{dR}\imag\left[\delta\Sigma_m \left(\frac{\delta
        v_{Rm}}{-\ii\sbar}\right)^*\right] \notag\\
  & = \frac{m}{2}\frac{dc_s^2}{dR}\imag\left[\ii \delta\Sigma_m \left(\frac{\delta
        v_{Rm}}{\sbar}\right)^*\right]
\end{align}
where we replaced the radial Lagrangian displacement by the Eulerian
radial velocity perturbation in Eq. \ref{baroclinic_torque}. 
The linearised momentum equations give
\begin{align}
  D\delta v_{Rm} =& 
  \ii\sbar\left[c_s^2\frac{d}{dR}\left(\frac{\delta\Sigma_m}{\Sigma}\right)
    + \frac{d}{dR}\delta\Phi_m\right] \notag\\ 
  &- \frac{2\ii
    m\Omega}{R}\left(c_s^2\frac{\delta\Sigma_m}{\Sigma} +
    \delta\Phi_m\right),
\end{align}
where $D\equiv \kappa^2 - \sbar^2$.

We now invoke local theory by setting $d/dR \to \ii k$ where $k$ is
real, and using the local solution to the Poisson equation
\begin{align}
  \delta\Phi_m = -\frac{2\pi G}{|k|}\delta\Sigma_m 
\end{align}
\citep{shu91}. Then the radial velocity perturbation becomes
\begin{align}
  \delta v_{Rm} =
  \frac{1}{D}\frac{\delta\Sigma_m}{\Sigma}\left(\frac{2\pi G
      \Sigma}{|k|} - c_s^2\right)\left(k\sbar + \frac{2\ii
      m\Omega}{R}\right),
\end{align}
so that 
\begin{align}
  \ii\delta\Sigma_m \left(\frac{\delta v_{Rm}}{\sbar}\right)^* =
  \frac{1}{D^*}\frac{|\delta\Sigma_m|^2}{\Sigma}\left(\frac{2\pi G
      \Sigma}{|k|} - c_s^2\right)\left(\ii k  + \frac{2
      m\Omega}{R\sbar^*}\right). 
\end{align}
For a neutral mode, $\sbar$, and therefore $D$ is real. Then, after
taking the imaginary part of the above expression, the torque
density is 
\begin{align}\label{tbg_explicit}
  T_\mathrm{BG} &=
  \frac{m}{2}\frac{dc_s^2}{dR}\left[\frac{k}{D}\frac{|\delta\Sigma_m|^2}{\Sigma}\left(\frac{2\pi G
      \Sigma}{|k|} - c_s^2\right)\right].
% &\simeq \frac{m}{2}\frac{dc_s^2}{dR}\left[\frac{k}{\left(\kappa^2 -
%         m^2\Omega^2\right)}\frac{|\delta\Sigma_m|^2}{\Sigma}\left(\frac{2\pi
%       G \Sigma}{|k|} - c_s^2\right)\right]. 
\end{align}
%where the second line follows from the low-frequency approximation
%$|\sigma|\ll m\Omega$. 
% To determine the sign of $T_\mathrm{BG}$ for low frequency one-arm spirals,
% we examine each factor separately below for $m=1$:
% \begin{itemize}
% \item We observe trailing spirals in our simulations, so $k>0$. 
% \item The factor $\kappa^2 - \Omega^2>0$ for our self-gravitating disc
%   models in the region where the spiral appears. 
% \item We measure $|k|\sim \pi G\Sigma/c_s^2$, so the last factor in
%   the square brackets is positive. Note that this factor is positive
%   provided $|k| < 2/HQ$ or wavelengths $\lambda > \pi Q H$.  
% \end{itemize}
We now use the local dispersion relation,
Eq. \ref{dispersion}, to make the replacement $D = 2\pi G \Sigma |k| -
k^2c_s^2$ in the above expression. This gives
\begin{align}\label{tbg_simple}
  T_\mathrm{BG} =
  \frac{m}{2}\frac{dc_s^2}{dR}\frac{1}{k}\frac{|\delta\Sigma_m|^2}{\Sigma}.  
\end{align} 
This torque density is negative for trailing waves in discs with
temperature decreasing outwards ($k>0$ and $dc_s^2/dR<0$,
respectively). Note that this conclusion does not rely on the
low-frequency approximation and holds for any $m$.

However, if the linear disturbance under consideration \emph{is} a
low-frequency $m=1$ 
trailing spiral wave, then it has negative angular
momentum. If $dc_s^2/dR<0$, as is typical for circumstellar or
protoplanetary discs, then $T_\mathrm{BG}<0$, and the background disc
applies a negative torque on the disturbance, which further decreases
its angular momentum. This suggests the spiral amplitude will grow.

Using $\jlin$ and $T_\mathrm{BG}$, we can obtain a rough estimate of
the growth rate $\gamma$ of linear perturbations as  
\begin{align}
  2\gamma \sim \frac{T_\mathrm{BG}}{\jlin},
\end{align}
where the factor of two accounts for the fact that the angular momentum
density is quadradtic in the linear perturbations. Inserting the above
expressions for $\jlin$ and $T_\mathrm{BG}$ for $m=1$ gives
\begin{align}
  2\gamma \sim
  -\frac{dc_s^2}{dR}\frac{1}{k\Sigma^2\Omega}\frac{|\delta\Sigma_1|^2}{|\xi_R
    + \ii\xi_\phi|^2}.
\end{align}
Now, using $\delta\Sigma_m/\Sigma = -\ii k \xi_R - \ii m \xi_\phi/R$
in the local approximation, we have
\begin{align}
  2\gamma \sim -\frac{dc_s^2}{dR}\frac{k}{\Omega}\frac{|\xi_R +
    \xi_\phi/kR|^2}{|\xi_R + \ii \xi_\phi|^2}
  \lesssim -\frac{dc_s^2}{dR}\frac{k}{\Omega}.
\end{align}
where we used $dc_s^2/dR <0$, $k>0$ and $|kR|\gg 1$ to obtain the
inequality. For $c_s^2 = c_{s0}^2 (R/R_0)^{-q}$ adopted in our disc
models,
\begin{align}\label{theoretical_rate}
  2\gamma \lesssim q\frac{c_s^2}{R}\frac{k}{\Omega}. 
\end{align}
Now we consider perturbations in a self-gravitating disc with characteristic
wavenumber $k \sim \pi G \Sigma/c_s^2 \sim \Omega/Q c_s$ (as seen in
our simulations). Then
\begin{align}
  \gamma \lesssim \frac{qh}{2Q}\Omega,
\end{align}
where we used $c_s/R\sim h\Omega$. For our fiducial disc models with
$q=1$, $Q=O(1)$, $h=0.05$, this gives $\gamma/\Omega < O(10^{-2})$,
consistent with simulation results.

Of course, we assumed a neutral mode to obtain expressions
for $\jlin$ and $T_\mathrm{BG}$ in the first place, so
Eq. \ref{theoretical_rate} is only an order-of-magnitude
constraint. Nevertheless, the expected growth rate is in reasonable
agreement with numerical results.

>>>>

\subsection{Unstable interaction between low-frequency $m=1$ spirals
  and the background disc due to an imposed temperature gradient}
The $m=1$ spiral that emerges in our simulations have low frequency
($\Omega_p < \Omega$),
grows slowly ($\gamma\ll\Omega_p$), and its spatial form is consistent local theory  
(\S\ref{fargo_m1}). Then to a first approximation, we can regard it 
as a neutral, low-frequency $m=1$ mode in satifying the 
dispersion relation Eq. \ref{dispersion} in local theory.  %attribute
                                %growth to temp gradient 
To demonstrate possible instability, %we 
% first construct a neutral low-frequency mode using local theory, 
we show that the torque density acting on such a mode due to the background
temperature gradient is negative, which enforces the mode, because it
has negative angular mometum.  
% Note that
% we will not be assuming a self-gravitating disc until the end of this
% section.

We begin by recalling the expression for the torque density due to a
locally isothermal equation of state,
\begin{align*}
  T_\mathrm{BG} = - \frac{m}{2}\frac{dc_s^2}{dR}\imag\left(\delta\Sigma_m\xi_R^*\right).   
\end{align*}
The Eulerian surface density perturbation is given by
\begin{align}\label{den_pert}
  \delta\Sigma_m = -\nabla\cdot\left(\Sigma\bm{\xi}\right) 
  = -\frac{1}{R}\frac{d}{dR}\left(R\Sigma \xi_R\right) - \frac{\ii m}{R}\Sigma\xi_\phi.  
\end{align}
We now invoke local theory by setting $d/dR \to \ii k$ where $k$ is
real, and assume $|kR|\gg m$ so that the second term on the right hand
side of Eq. \ref{den_pert} is negligible. Then    
\begin{align}
  \delta\Sigma _m  \simeq -\ii k \Sigma \xi_R,
\end{align}
which gives
\begin{align}
  T_\mathrm{BG} = \frac{m}{2}\frac{dc_s^2}{dR}k\Sigma |\xi_R|^2. 
\end{align}
% \begin{align}
%   T_\mathrm{BG} &= -\frac{m}{2}\frac{dc_s^2}{dR}\imag\left[\delta\Sigma_m \left(\frac{\delta
%         v_{Rm}}{-\ii\sbar}\right)^*\right] \notag\\
%   & = \frac{m}{2}\frac{dc_s^2}{dR}\imag\left[\ii \delta\Sigma_m \left(\frac{\delta
%         v_{Rm}}{\sbar}\right)^*\right]
% \end{align}
% where we replaced the radial Lagrangian displacement by the Eulerian
% radial velocity perturbation in Eq. \ref{baroclinic_torque}. 
% The linearised momentum equations give
% \begin{align}
%   D\delta v_{Rm} =& 
%   \ii\sbar\left[c_s^2\frac{d}{dR}\left(\frac{\delta\Sigma_m}{\Sigma}\right)
%     + \frac{d}{dR}\delta\Phi_m\right] \notag\\ 
%   &- \frac{2\ii
%     m\Omega}{R}\left(c_s^2\frac{\delta\Sigma_m}{\Sigma} +
%     \delta\Phi_m\right),
% \end{align}
% where $D\equiv \kappa^2 - \sbar^2$. 
% We now invoke local theory by setting $d/dR \to \ii k$ where $k$ is
% real, and using the local solution to the Poisson equation
% \begin{align}
%   \delta\Phi_m = -\frac{2\pi G}{|k|}\delta\Sigma_m 
% \end{align}
% \citep{shu91}. Then the radial velocity perturbation becomes
% \begin{align}
%   \delta v_{Rm} =
%   \frac{1}{D}\frac{\delta\Sigma_m}{\Sigma}\left(\frac{2\pi G
%       \Sigma}{|k|} - c_s^2\right)\left(k\sbar + \frac{2\ii
%       m\Omega}{R}\right),
% \end{align}
% so that 
% \begin{align}
%   \ii\delta\Sigma_m \left(\frac{\delta v_{Rm}}{\sbar}\right)^* =
%   \frac{1}{D^*}\frac{|\delta\Sigma_m|^2}{\Sigma}\left(\frac{2\pi G
%       \Sigma}{|k|} - c_s^2\right)\left(\ii k  + \frac{2
%       m\Omega}{R\sbar^*}\right). 
% \end{align}
% For a neutral mode, $\sbar$, and therefore $D$ is real. Then, after
% taking the imaginary part of the above expression, the torque
% density is 
% \begin{align}\label{tbg_explicit}
%   T_\mathrm{BG} &=
%   \frac{m}{2}\frac{dc_s^2}{dR}\left[\frac{k}{D}\frac{|\delta\Sigma_m|^2}{\Sigma}\left(\frac{2\pi G
%       \Sigma}{|k|} - c_s^2\right)\right].
% % &\simeq \frac{m}{2}\frac{dc_s^2}{dR}\left[\frac{k}{\left(\kappa^2 -
% %         m^2\Omega^2\right)}\frac{|\delta\Sigma_m|^2}{\Sigma}\left(\frac{2\pi
% %       G \Sigma}{|k|} - c_s^2\right)\right]. 
% \end{align}
% We now use the local dispersion relation,
% Eq. \ref{dispersion}, to make the replacement $D = 2\pi G \Sigma |k| -
% k^2c_s^2$ in the above expression. This gives
% \begin{align}\label{tbg_simple}
%   T_\mathrm{BG} =
%   \frac{m}{2}\frac{dc_s^2}{dR}\frac{1}{k}\frac{|\delta\Sigma_m|^2}{\Sigma}.  
% \end{align}  
This torque density is negative for trailing waves in discs with
a fixed temperature profile decreasing outwards ($k>0$ and $dc_s^2/dR<0$,
respectively). Note that this conclusion does not rely on the
low-frequency approximation and holds for any $m$.

However, if the linear disturbance under consideration \emph{is} a
low-frequency $m=1$ mode, then it has negative angular
momentum. If it is a trailing spiral and $dc_s^2/dR<0$, as is typical
for circumstellar or protoplanetary discs, then $T_\mathrm{BG}<0$, and
the background disc applies a negative torque on the disturbance,
which further decreases its angular momentum. This suggests the spiral
amplitude will grow.

Using $\jlin$ and $T_\mathrm{BG}$, we can obtain a rough estimate of
the growth rate $\gamma$ of linear perturbations due to the background
torque as  
\begin{align}
  2\gamma \sim \frac{T_\mathrm{BG}}{\jlin},
\end{align}
where the factor of two accounts for the fact that the angular momentum
density is quadradtic in the linear perturbations. Inserting the above
expressions for $\jlin$ and $T_\mathrm{BG}$ for $m=1$ gives
\begin{align}
  2\gamma \sim
  -\frac{dc_s^2}{dR}\frac{k}{\Omega}\frac{|\xi_R|^2}{|\xi_R + \ii
    \xi_\phi|^2}. 
%  \lesssim -\frac{dc_s^2}{dR}\frac{k}{\Omega}.
\end{align}
In Appendix \ref{horizontal_displacements}, we argue $\xi_\phi \simeq
2\ii \xi_R$ for the modes of interest. Then 
%To further simplify this expression, we need to relate $\xi_R$ and
%$\xi_\phi$. 
% \begin{align}
%   2\gamma \sim
%   -\frac{dc_s^2}{dR}\frac{1}{k\Sigma^2\Omega}\frac{|\delta\Sigma_1|^2}{|\xi_R
%     + \ii\xi_\phi|^2}.
% \end{align}
% Now, using $\delta\Sigma_m/\Sigma = -\ii k \xi_R - \ii m \xi_\phi/R$
% in the local approximation, we have
% \begin{align}
%   2\gamma \sim -\frac{dc_s^2}{dR}\frac{k}{\Omega}\frac{|\xi_R +
%     \xi_\phi/kR|^2}{|\xi_R + \ii \xi_\phi|^2}
%   \lesssim -\frac{dc_s^2}{dR}\frac{k}{\Omega}.
% \end{align}
% where we used $dc_s^2/dR <0$, $k>0$ and $|kR|\gg 1$ to obtain the
% inequality. 
%where we used $dc_s^2/dR <0$ and $k>0$ to obtain the inequality. 
for the temperature profiles $c_s^2 = c_{s0}^2 (R/R_0)^{-q}$ as
adopted in our disc models,  
\begin{align}
  2\gamma \sim q\frac{c_s^2}{R}\frac{k}{\Omega}. 
\end{align}
For perturbations in a self-gravitating disc with characteristic
wavenumber $k \sim \pi G \Sigma/c_s^2 \sim \Omega/Q c_s$, as seen in
our simulations, we have 
\begin{align}\label{theoretical_rate}
  \gamma \sim \frac{qh}{2Q}\Omega,
\end{align}
where we used $c_s/R\sim h\Omega$. For our fiducial disc models with
$q=1$, $Q=O(1)$, $h=0.05$, this gives $\gamma/\Omega \sim O(10^{-2})$,
consistent with simulation results.

%Of course, we assumed a neutral mode to obtain expressions
%for $\jlin$ and $T_\mathrm{BG}$ in the first place, 
Note that by using local theory we have implicitly
assumed a neutral mode because the dispersion relation,
Eq. \ref{dispersion}, states that non-axisymmetric modes are
stable for real $k$. Eq. \ref{theoretical_rate} should be interpreted
as an order-of-magnitude estimate of the perturbative effect of a
temperature gradient on local, low-frequency $m=1$ modes.

%neutral modes are not actually neutral)
%gamma << omega_p << omega 
%so
%Eq. \ref{theoretical_rate} is only an order-of-magnitude
%constraint. Nevertheless, the expected growth rate is in reasonable
%agreement with numerical results.

### 2014-11-15 paper/summary.tex
update
simplified discussion

%non-sg sims, Be stars, eccentric discs
We stress that destabilisation through the temperature gradient
does not explicitly depend on self-gravity. In principle, this effect
will destabilise low-frequency $m=1$ trailing spirals regardless of
its origin. In our models, self-gravity coupled with the disc
structure permits a neutral one-arm mode, which is then destabilised. 
It would be of interest to explore other origins of the neutral
one-arm spiral. One possibility is in Be star discs
\citep{rivinius13}, for which 
one-arm oscillations may explain long-timescale variations in their
emission lines \citep[see e.g.][and references
therein]{okasaki97,papaloizou06c,ogilvie08}. These studies employ
strictly isothermal disc models, so it would be interesting to explore
the effect of a temperature gradient on the stability of these
oscillations.

>>>>

%non-sg sims, Be stars, eccentric discs
% We stress that destabilisation through the temperature gradient
% does not explicitly depend on self-gravity. In principle, this effect
% will destabilise low-frequency $m=1$ trailing spirals regardless of
% its origin.
In our models and interpretation, self-gravity coupled with the disc 
structure permits a neutral one-arm mode, which is then destabilised.   
It would be interesting to explore other origins of the neutral
one-arm spiral. One possibility is in Be star discs
\citep{rivinius13}, for which 
one-arm oscillations may explain long-timescale variations in their
emission lines \citep[see e.g.][and references
therein]{okasaki97,papaloizou06c,ogilvie08}. These studies employ
strictly isothermal disc models, so it would be interesting to explore
the effect of a temperature gradient on the stability of these
oscillations.

%vertical shear instability?
